\documentclass{assets/fieldnotes}

\title{Kono (Sierra Leone)}
\author{LING3020/5020}
\date{University of Pennsylvania, Spring 2025\\01/29/2025 Phonetics and Phonology I }

\setcounter{secnumdepth}{4} %enable \paragraph -- for subsubsubsections

\begin{document}

\maketitle
\tableofcontents

\newpage 
%(Daniel absent today)



%dipthongs / vowel sequences 
%vowel nasalization - CVN/CVVN (compared to CV/CVV) 
\section{Consonants - stop voicing and place contrast -- Alex}
\jf{Needs Transcription}\\


\section{/gb/, /kp/  versus /b/, /p/ --- Wesley}

\wml{Planned minimal pairs -- I will try at least a few pairs for each consonantal contrast, so that we can get tokens for a few different phonological environments.}\\
\\
\wml{\textipa{/\t{gb}/ vs. /b/}}\\

\exg. \textipa{b\'{e}k\`{i}}\\
bad\\
`bag'

\exg. \textipa{\texttoptiebar{gb}\'{e}k\`{i}}\\
Kono.surname\\
`(a Kono surname)'

\exg. \textipa{\t{gb}\^{i}}\\
pumpkin\\
`pumpkin'

\exg. \textipa{b\^{i}}\\
termite\\
`termite'

\exg. \textipa{b\'{i}}\\
today\\
`today'

\exg. \textipa{\t{gb}o\textlengthmark}\\
body\\
`body'

\exg. \textipa{\t{gb}\`o\textlengthmark}\\
padlock\\
`padlock'

\exg. \textipa{\t{gb}\^o\textlengthmark}\\
curse\\
`curse'

\exg. \textipa{\t{gb}\'{o}\textlengthmark}\\
book\\
`book'

\exg. \textipa{bo\textlengthmark}\\
hand\\
`hand'

\exg.\textipa{\t{gb}\'{e}}\\
all\\
`all'

\exg. \textipa{b\`{e}}\\
to give\\
`to give'

\exg. \textipa{\t{gb}a}\\
dry\\
`dry'

\exg. \textipa{ba}\\
big\\
`big'

\exg. \textipa{sam\t{gb}a}\\
drum\\
`drum'

\exg. \textipa{samba}\\
basket\\
`basket'

\exg. \textipa{jamba}\\
leaf\\
`leaf'

\wml{\textipa{/\t{gb}/} vs. \textipa{/\t{kp}/}}\\



\wml{I had one token of \textipa{\t{kp}} in my January 27 notes and wasn't sure if this was truly a voiceless phoneme. Foday-Ngongou reports interspeaker phonological variation in voiced and voiceless counterparts.}

\exg. \textipa{\t{kp}ajne}\\
to fish\\
`to fish'

\wml{There's also...:}\\

\exg. \textipa{\t{kp}am\t{gb}a}\\
oven\\
`oven'

\wml{(Near)-minimal pairs:}

\exg. \textipa{\t{kp}e\t{kp}e}\\
frog\\
`frog'

\exg. \textipa{\t{kp}\textepsilon\t{kp}\textepsilon}\\
comb\\
`comb'

\exg. \textipa{\t{gb}\textepsilon\t{gb}\textepsilon}\\
stubborn\\
`stubborn'

\exg. \textipa{\t{gb}e\ng\t{gb}\textepsilon}\\
boat\\
`boat'

\wml{\textipa{/p/} vs. \textipa{/b/}}

\exg. \textipa{p\textsuperscript h u}\\
overseas\\
`overseas'

\exg. \textipa{b\'{u}}\\
stomach\\
`stomach'

\exg. \textipa{\t{gb}u}\\
swollen\\
`swollen'

\section{lateral, approximates -- Chun-Hung}

\chs{/j/ vs. /w/}

\exg. ju \\ 
rope\\
`rope' (m83)

\exg. wu \\
you (pl)\\
`you (pl)' (m83)

\exg. wuu \\
dog\\
`dog'

\exg. ja \\
eye\\
`eye'

\exg. jaa \\
peanut\\
`peanut'

\exg. jaa \\
lion\\
`lion' (m83)

\exg. wa \\
big\\
`big'

\exg. jai \\
name of a person\\
`name of a person' (n85)

\exg. wai \\
resting place\\
`resting place' (n85)

\exg. ji \\
water\\
`water'

\exg. wi \\
blood\\
`blood' (F85)

\chs{/l/ and /\textipa{\*r}/}

\chs{/l/ is restricted to the word-medial position according to F85, unless loanwords.} 

\chs{/\textipa{\*r}/ is not documented}

\exg. dulu \\
five\\
`five'

\exg. dusu \\
fiber from bamboo tree\\
`fiber from bamboo tree' (m83)

\exg. w\textipa{O}l\textipa{O} \\
six\\
`six'

\exg. d\textipa{O}f\textipa{O} \\
insignificant\\
`insignificant'

\exg. \textipa{\*r}umu \\
room\\
`room'

\exg. mumun\textipa{E} \\
a dumb person\\
`a dumb person' (n85)

\section{Back vowels - height contrast --- Joey}

\jf{Short Vowels:}

\exg. tù \\
to pound/hit/touch\\
`to pound/hit/touch'

\exg. tù \\
oil\\
`oil'

\exg. tó \\
to stay/remain\\
`to stay/remain' 
\jf{This works better in the context of 'stay here/don't follow me.' Anthony said that it usually needs more context tó ne = 'stay'}

\exg. siné \\
to stay/remain\\
`to stay/remain (seated here)'

\exg. nɔ́ \\
there\\
`there' ('not here') 
\jf{This can't stand alone. It works better in the context of 'we are there.' - bé nɔ́}

\exg. pandɛ \\
there\\
`there' ('not here')

\exg. bù̆ \\
stomach\\
`Stomach'

\exg. bɔ́ \\
to go out\\
`to go out' (used if you're telling someone to go as well it is more common in the form of bɔ́wa)

\exg. dú \\
to share/distribute\\
`to share/distribute'

\exg. dòmâ \\
shirt\\
`shirt'

\exg. dɔ̀ndɔ̀ \\
one\\
`one' (when ennumerating things, not when counting)

\exg. dòmâdɔ̀ndɔ̀wɛ \\
shirt one\\
`one shirt'

\exg. t͡ʃɛlɛ \\
one\\
`one' (when counting, not when ennumerating things)


\jf{Long Vowels:}


\exg. kôː \\
to fatten\\
`To fatten (an animal)'

\ex. túː \\
`To become fat'\\
\jf{Anthony mentioned that this actually means the fat that is inside of an animal}

\ex. tóː \\
`To envy/hate/be jealous of'

\ex. daŋba\\
`Dirty'

\ex. bóː\\
`Body'

\ex. gbôː\\
`Hand'





\section{Affricates and fricatives -- Jan}

\ex. \textipa{s\'EI} \\
`to climb' % originally [se]

\ex. \textipa{\texttoptiebar{tS}\'En\'E} \\
`leg' 

\ex. \textipa{\texttoptiebar{pf}uu} \\ 
`overseas, abroad' 

\ex. \textipa{mako\`a} \\
`brush'

\ex. \textipa{\texttoptiebar{mb}ENamak\^of\'en\`e} \\
`thing used to clean spoon'

\ex. \textipa{k\`o\`o} \\
`wash'

\ex. \textipa{cE} \\
`to do'

\ex. \textipa{c\`E\`E} \\
`to call'

\ex. \textipa{\texttoptiebar{mb}\^acen\`e} \\
`neighbor'

\ex. \textipa{cim\'a\`a} \\
`cold'

\ex. \textipa{\texttoptiebar{gb}ak\`uo} \\
`adult', `older person'

\ex. \textipa{musukaku\`o} \\
`adult man'

\ex. \textipa{\texttoptiebar{gb}akokaj} \\
`adult man'

\ex. \textipa{mukuM\texttoptiebar{gb}abuku} \\
`adult man'

\ex. \textipa{\texttoptiebar{gb}\'e\'e} \\
`meet'

\ex. \textipa{\texttoptiebar{gb}\'en\`e} \\
`meet'

\ex. \textipa{sw\'e\`e} \\
`meat'

\ex. \textipa{sw\`E\`E} \\
`bean'

\ex. \textipa{eab\'asi} \\
`onion'

\ex. \textipa{cEnJEnE} \\
`football'

\ex. \textipa{sw\'in\`e} \\
`rest'

\ex. \textipa{cEne} \\
`house'


\section{Nasal consonants - place contrasts Lex} 

\ex. koŋg wa or kon tʃenama or kom ba \\
`Big tree'

\ex.koné \\
`mountain'

\ex. koˈŋgʷé tʃenama, koˈŋgʷé wa \\
`mountain'

\ex. ɲinba\\
`big tooth'

\ex. ɲinɛ\\
`tooth'

\ex. nin-ba\\
`big tooth'

\ex. n-ɲinɛ\\
`my tooth'

\ex. musú\\
`woman'

\ex. mɔɛ̃ \\
`human'

\ex.  mɔɛ̃-ɡ͡bu \\
`human being'

\ex. mɔɛ̃kama\\
`human man'

\ex. wiinE\\
`deer'

\ex. senɛ\\
`stone'

\ex. sɛnɛ\\
`rice farm'

\ex. bùːbùː\\
`vegetable farm'

\ex. yadí\\
`farm of cash crops'

\ex. nɛːnɛnɛ\\
`tongue'

\ex. n-nɛːnɛnɛ\\
`my tongue'

\section{Front vowels - height contrast -- Giang}

\wml{Minimal pairs we got from last time:}\\

\ex. \textipa{tS\'en\^E}\\
`house'
\g{The falling tone on the second syllable might just be an end-of-word thing...}\\

\ex. \textipa{tS\`en\'E}\\
`leg'

\ex. \textipa{f\'in\`e}\\
`black'
\g{Foday-Ngongou has this as \textipa{f\'in\^E}.}\\

\ex. \textipa{f\'en\`e}\\
`thing'
\g{I think the contrast between black and thing is just the first vowel: i vs e.}\\

Other words for front vowels:
\ex. \textipa{b\`i}\\
`today'


\ex. \textipa{b\`e:}\\
`redemption fee (in a secret society)'

\ex. \textipa{b\`e}\\
`to give'
\g{Near minimal set...}\\

\ex. \textipa{d\'i}\\
`sweet'

\ex. \textipa{d\1e}\\
`mother'

\ex. \textipa{d\`i:}\\
`to get used to'
\g{Near minimal set}\\

\ex. \textipa{n\`E}\\
`here'

\ex. \textipa{m\'E}\\
`to stay longer, live longer, overstay'

\ex. \textipa{f\`i:}\\
`to miss, mistake, e.g: a person; a type of rice'

\ex. \textipa{f\'E}\\
`to blow, inflate; seize'

\ex. \textipa{\t{kp}\`E}\\
`to fry; difficult, hard'

\ex. \textipa{\t{kp}\'e:\t{kp}\^e}\\
`curse'
\g{Near minimal pair with ``to fry; difficult''}\\

\ex. \textipa{s\`E}\\
`to climb'

\ex. \textipa{s\`ic\`i}\\
`an inner feeling, suspicion'

\ex. \textipa{s\`Ec\`E}\\
`a type of musical instrument'
\g{This is a minimal pair but the meaning is so weird...}\\
\jal{he said this meant ``this s\textipa{EE}, it's bimorphemic"}

\ex. \textipa{c\'ec\^e}\\
`ringworm'

\ex. \textipa{c\'ic\^i:}\\
`a round hut'
\g{This is again a minimal pair with a weird meaning...}\\
\jal{what's weird about the meaning?}

\ex. \textipa{f\'if\^i:}\\
`mud, sand'

\ex. \textipa{f\'Ef\^E:}\\
`a spray of rain'
\g{This is again a minimal pair with a weird meaning...}

\section{Vowel Length Contrast - Mingyang}
\begin{itemize}
    \item \textbf{i vs. i:}
    \ex. si:\\
        `music'

    \ex. si\\
        `to sit'
    
    \ex. donbi:\\
        `orange'

    \ex. bi\\
        `today'

    \item \textbf{a vs. a:}
    \ex. banfa:\\
        `pants'

    \ex. banfa\\
        `disrespect'

    \ex. fa\\
        `father'

    \ex. k$^w$a\\
        `monkey'
    
    \ex. ka:\\
        `snake'

    \ex. ta:\\
        `calabash'

    \ex. ta\\
        `fire'

    \item \textbf{ɛ vs. ɛ:}
    \ex. t͡ʃɛ\\
        `this'

    \ex. t͡ʃɛ:\\
        `war'

    \ex. tɛtɛ\\
        `spider'
    
    \ex. tɛ:\\
        `chicken'

    \ex. dɛ\\
        `mother'
        

    \ex. m$^w$ɛ\\
        `person'

    \ex. mɛ:\\
        `knife'

    \ex. ɲɛː\\
        `fish'

    \item \textbf{ɔ vs. ɔ:}
    \ex. bɔ\\
        `to comb'

    \ex. bɔ:\\
        `to curse'
    
    \item \textbf{e vs. e:}
    \ex. t$^h$e:\\
        `sun'

    \ex. te\\
        `to break'

    \item \textbf{u vs. u:}
    \ex. bu:\\
        `garden'

    \ex. bu\\
        `stomach'
    
    \ex. du:\\
        `to bend'

    \ex. du\\
        `town'

    \item \textbf{o vs. o:}
    \ex. bo:\\
        `hand'
        
    \ex. bo\\
        `to cut'
        
    \ex. to:\\
        `ear'

\end{itemize}

\end{document}
